---
title:  Homework 1
author: "[Jacob Carey](mailto:jcarey15@jhu.edu)"
date:   "`r Sys.Date()`"
output: pdf_document
---

```{r readin, echo=FALSE}
load("nmes.rdata")
library(ggplot2)
library(dplyr, quietly=TRUE)
library(tidyr)
knitr::opts_chunk$set(cache=TRUE)
```

Problem 1
=========

Model A

```{r p1-a}
n <- 250
B.0 <- -2
B.1 <- 2

set.seed(1337)
X <- runif(n, 0, 1)

logit <- B.0 + B.1 * X
y <- rbinom(n, 1, exp(logit)/(1+exp(logit)))

model.A <- data_frame(x=X, y=y, model="A")

model.A %>%
    ggplot(aes(x=x, y=y)) +
    geom_jitter(position=position_jitter(height=0.1)) +
    ggtitle("Model A")
```

Model B

```{r p1-b}
n <- 250
B.0 <- -2
B.1 <- 4

set.seed(1337)
X <- runif(n, 0, 1)

logit <- B.0 + B.1 * X
y <- rbinom(n, 1, exp(logit)/(1+exp(logit)))

model.B <- data_frame(x=X, y=y, model="B")

model.B %>%
    ggplot(aes(x=x, y=y)) +
    geom_jitter(position=position_jitter(height=0.1)) +
    ggtitle("Model B")
```

```{r p1-table}
regr <- function(data) {
    coef <- lm(log.odds~stratum, data=data)$coefficients
    return(cbind(data, data.frame(pred.log.odds=coef[1] + coef[2]*data$stratum)))
}

table <- bind_rows(model.A, model.B) %>%
    # create stratum column using cut function
    mutate(stratum=cut(x, 0:5/5, labels=0:4/5),
           stratum=as.numeric(as.character(stratum)),
           stratum=stratum + 0.1) %>%
    # group by stratum
    group_by(stratum, model) %>%
    # count rows and number y==1 in stratum
    summarise(n=n(),
              ones=sum(y == 1)) %>%
    # calculat p, odds, and log odds
    mutate(p=ones / n,
           odds=p / (1 - p),
           log.odds=log(odds)) %>%
    group_by(model) %>%
    do(regr(.)) %>%
    mutate(pred.p=1/(1+exp(-pred.log.odds))) %>%
    arrange(model)

knitr::kable(table, digits=2, col.names=c("Model",
                                          "Stratum of X",
                                          "n",
                                          "# Y=1",
                                          "p",
                                          "odds",
                                          "log odds",
                                          "prediction log odds",
                                          "predicted p"))

```

Logistic regression estimates the probability of an outcome, via the
logarithm odds.

Problem 2
=========

**1a**

```{r 2-1-a}
data <- nmes %>%
    tbl_df() %>%
    filter(eversmk != ".") %>%
    mutate(mscd=lc5+chd5>0,
           mscd=as.integer(mscd),
           eversmk=as.integer(eversmk)) %>%
    select(age=lastage, mscd, eversmk)

table <- data %>%
    count(mscd, eversmk) %>%
    spread(mscd, n)

knitr::kable(table, col.names=c("Ever smoke", "mscd=0", "mscd=1"))

table <- table[, 2:3]
odds.ratio <- table[1, 1] * table[2, 2] / (table[1, 2] * table[2, 1])[1, ]
log.or <- log(odds.ratio)
se.log.or <- sqrt(sum(1/table))
log.or.ci <- log.or + c(-1, 1) * qnorm(0.975) * se.log.or

```

We see a log odds ratio of `r round(log.or, 2)` with standard error
`r round(se.log.or, 2)` and 95% confidence interval 
`r round(log.or.ci, 2)`.

**1b**

```{r 2-1-b}
model <- glm(mscd ~ eversmk, data=data, family="binomial")
```

We see a regresion coefficient of `r round(coef(model)[2], 2)` and
a standard error of `r round(sqrt(diag(vcov(model))), 2)[2]` which are
similar to above.

**1c**

```{r 2-1-c}
model <- glm(eversmk ~ mscd, data=data, family="binomial")
```

We see a regresion coefficient of `r round(coef(model)[2], 2)` and
a standard error of `r round(sqrt(diag(vcov(model))), 2)[2]` which are
similar to above.

**1d**

The logistic regression coefficient estimates the increase in log odds
of the outcome per unit increase of the covariate. The invariance 
principle of the odds ratio indicates that odds ratio of an exposure
and outcome is the same for the reversed expression.

**2a**
```{r 2-2-a}
log.or <- function(mscd, eversmk) {
    tbl <- table(mscd, eversmk)
    or <- tbl[1, 1] * tbl[2, 2] / (tbl[1, 2] * tbl[2, 1])
    return(log(or))
}

var.log.or <- function(mscd, eversmk) {
    tbl <- table(mscd, eversmk)
    var.log.or <- sum(1/tbl)
    return(var.log.or)
}

table <- bind_rows(data %>%
    mutate(age.strat=cut(age, c(0, 50, 60, 70, 150),
                         labels=c("<50", "51-60", "61-70", "70+"))) %>%
    group_by(age.strat) %>%
    summarise(log.or=log.or(mscd, eversmk),
              se.log.or=sqrt(var.log.or(mscd, eversmk))) %>%
    mutate(inv.var=1/se.log.or^2,
           weight=inv.var/sum(inv.var)),
    data %>%
        mutate(age.strat="Ignoring Age") %>%
        group_by(age.strat) %>%
        summarise(log.or=log.or(mscd, eversmk),
                  se.log.or=sqrt(var.log.or(mscd, eversmk))) %>%
        mutate(inv.var=1/se.log.or^2,
               weight=inv.var/sum(inv.var)))

knitr::kable(table, digits=3, col.names=c("Age Stratum",
                                          "Log OR",
                                          "Std Error",
                                          "1/Variance",
                                          "Weight"))
```

**2b**
```{r 2-2-b}
table <- table %>% filter(age.strat != "Ignoring Age")
weighted.log.or <- weighted.mean(table$log.or, table$weight)
weighted.se <- sqrt(1/sum(table$inv.var))
```

We calculate the weighted log or as `r round(weighted.log.or, 2)` and
the standard error as `r round(weighted.se, 2)`.

**2c**
```{r 2-2-c}
model <- data %>%
    mutate(age.strat=cut(age, c(0, 50, 60, 70, 150),
                         labels=c("<50", "51-60", "61-70", "70+"))) %>%
    glm(mscd ~ eversmk + age.strat, family="binomial", data=.)
```

Using logistic regression, we calculate the coefficient and standard
error as `r round(coef(model)[2], 3)` and 
`r round(sqrt(diag(vcov(model))), 3)[2]` respectively.

**2d**
```{r 2-2-d}
model <- data %>%
    mutate(age.center=age-60,
           age.plus=ifelse(age.center>0, age.center, 0),
           age.cube=age.plus^3) %>%
    glm(mscd~eversmk+age.center+age.plus+age.cube, 
        family="binomial", data=.)
```

Repeating the analysis , we calculate the coefficient and standard
error as `r round(coef(model)[2], 3)` and 
`r round(sqrt(diag(vcov(model))), 3)[2]` respectively.

**2e**
```{r 2-2-e}
model <- data %>%
    mutate(age.strat=cut(age, c(0, 50, 60, 70, 150),
                         labels=c("<50", "51-60", "61-70", "70+"))) %>%
    glm(eversmk ~ mscd + age.strat, family="binomial", data=.)
```

Using logistic regression, we calculate the coefficient and standard
error as `r round(coef(model)[2], 3)` and 
`r round(sqrt(diag(vcov(model))), 3)[2]` respectively. These numbers are
fairly close to the estimates in 2d and are the same as 2c.

**2f**
```{r 2-2-f}
model <- data %>%
    mutate(age.center=age-60,
           age.plus=ifelse(age.center>0, age.center, 0),
           age.cube=age.plus^3) %>%
    glm(mscd~eversmk+age.center+age.plus+age.cube, 
        family="binomial", data=.)

predicted <- data %>%
    mutate(age.center=age-60,
           age.plus=ifelse(age.center>0, age.center, 0),
           age.cube=age.plus^3) %>%
    predict(model, .)

data$predicted <- 1/(1+exp(-predicted))

ggplot(data, aes(x=age, y=mscd)) +
    geom_jitter(aes(colour=factor(eversmk)),
                position=position_jitter(height=0.2),
                alpha=0.4) +
    stat_smooth(aes(colour=factor(eversmk))) + 
    geom_line(aes(y=predicted, colour=factor(eversmk)),
              linetype="dashed")
```

There does not appear to be effect modification of the smoking-mscd 
association by age based on this graph.

**2g**
A proposed extended model might also include an interaction term for
eversmk.
```{r 2-2-g}
ext.model <- data %>%
    mutate(age.center=age-60,
           age.plus=ifelse(age.center>0, age.center, 0),
           age.cube=age.plus^3) %>%
    glm(mscd~eversmk*(age.center+age.plus+age.cube),
        family="binomial", data=.)

p.value <- 1 - pchisq(ext.model$deviance - model$deviance,
                      model$df.residual - ext.model$df.residual)
```

The p value from the likelihood ratio test is `r p.value`, indicating that
the extended model is not better.

