---
title:  Homework 2
author: "[Jacob Carey](mailto:jcarey15@jhu.edu)"
date:   "`r Sys.Date()`"
output: pdf_document
---

```{r readin, echo=FALSE}
knitr::opts_chunk$set(cache=TRUE)
load("../data/nmes.rdata")
library(ggplot2)
library(dplyr, quietly=TRUE)
library(tidyr)
library(pROC)
```

Problem 1
=========
```{r p1}
data <- nmes %>%
    filter(lastage >= 40) %>%
    mutate(large.exp=as.integer(totalexp > 2e4)) %>%
    select(large.exp, lc5, chd5, educate, race=RACE3,
           male, age=lastage, region=sregion, marital,
           poverty=povstalb)

model.1 <- glm(large.exp~lc5 + chd5, family="binomial", data=data)
model.2 <- glm(large.exp~lc5 + chd5 + race + male, 
               family="binomial", data=data)
model.3 <- glm(large.exp~lc5 + chd5 + race + male + age, 
               family="binomial", data=data)
model.4 <- glm(large.exp~lc5 + chd5 + race + male * age, 
               family="binomial", data=data)
model.5 <- glm(large.exp~lc5 + chd5 + race + male * age + factor(poverty),
               family="binomial", data=data)
model.6 <- glm(large.exp~lc5 + chd5 + race + male * age + factor(marital),
               family="binomial", data=data)

p.value.1.2 <- 1 - pchisq(model.1$deviance - model.2$deviance,
                          model.1$df.residual - model.2$df.residual)
p.value.1.3 <- 1 - pchisq(model.1$deviance - model.3$deviance,
                          model.1$df.residual - model.3$df.residual)
p.value.3.4 <- 1 - pchisq(model.3$deviance - model.4$deviance,
                          model.3$df.residual - model.4$df.residual)
p.value.4.5 <- 1 - pchisq(model.4$deviance - model.5$deviance,
                          model.4$df.residual - model.5$df.residual)
p.value.4.6 <- 1 - pchisq(model.4$deviance - model.6$deviance,
                          model.4$df.residual - model.6$df.residual)
```

Using the likelihood ratio test we calculate the following p values:  
Comparing model 1 to 2: `r round(p.value.1.2, 4)`  
Comparing model 1 to 3: `r round(p.value.1.3, 4)`  
Comparing model 3 to 4: `r round(p.value.3.4, 4)`  
Comparing model 4 to 5: `r round(p.value.4.5, 4)`  
Comparing model 4 to 6: `r round(p.value.4.6, 4)`  

Problem 2
=========
Based on these calculations, I decided to use model 6, which include
lc5, chd5, race, male, age, an interaction between male and age, and
marital status as covariates.

I estimated the following $\beta$ coefficients:
`r coef(model.6)`.

Problem 3
=========
```{r p3}
data <- data %>%
    mutate(prob=predict(model.6, type=c("response")))

roc <- roc(large.exp~prob, data=data)
plot(roc, legacy.axes=TRUE)
```
