---
title:  Homework 2
author: "[Jacob Carey](mailto:jcarey15@jhu.edu)"
date:   "`r Sys.Date()`"
output: pdf_document
---

```{r chunks, echo=FALSE}
knitr::opts_chunk$set(cache=TRUE,
                      fig.width=6,
                      fig.height=4)
```


```{r readin, echo=FALSE}
load("../data/nmes.rdata")
library(ggplot2)
library(dplyr)
library(tidyr)
library(pROC)
theme_set(theme_classic())
```

Problem 1
=========
```{r p1}
data <- nmes %>%
    filter(lastage >= 40) %>%
    mutate(large.exp=as.integer(totalexp > 2e4)) %>%
    select(large.exp, lc5, chd5, educate, race=RACE3,
           male, age=lastage, region=sregion, marital,
           poverty=povstalb)

model.1 <- glm(large.exp~lc5 + chd5, family="binomial", data=data)
model.2 <- glm(large.exp~lc5 + chd5 + race + male, 
               family="binomial", data=data)
model.3 <- glm(large.exp~lc5 + chd5 + race + male + age, 
               family="binomial", data=data)
model.4 <- glm(large.exp~lc5 + chd5 + race + male * age, 
               family="binomial", data=data)
model.5 <- glm(large.exp~lc5 + chd5 + race + male * age + factor(poverty),
               family="binomial", data=data)
model.6 <- glm(large.exp~lc5 + chd5 + race + male * age + factor(marital),
               family="binomial", data=data)

p.value.1.2 <- 1 - pchisq(model.1$deviance - model.2$deviance,
                          model.1$df.residual - model.2$df.residual)
p.value.1.3 <- 1 - pchisq(model.1$deviance - model.3$deviance,
                          model.1$df.residual - model.3$df.residual)
p.value.3.4 <- 1 - pchisq(model.3$deviance - model.4$deviance,
                          model.3$df.residual - model.4$df.residual)
p.value.4.5 <- 1 - pchisq(model.4$deviance - model.5$deviance,
                          model.4$df.residual - model.5$df.residual)
p.value.4.6 <- 1 - pchisq(model.4$deviance - model.6$deviance,
                          model.4$df.residual - model.6$df.residual)
```

Using the likelihood ratio test we calculate the following p values:  
Comparing model 1 to 2: `r round(p.value.1.2, 4)`  
Comparing model 1 to 3: `r round(p.value.1.3, 4)`  
Comparing model 3 to 4: `r round(p.value.3.4, 4)`  
Comparing model 4 to 5: `r round(p.value.4.5, 4)`  
Comparing model 4 to 6: `r round(p.value.4.6, 4)`  

Problem 2
=========
Based on these calculations, I decided to use model 6, which include
lc5, chd5, race, male, age, an interaction between male and age, and
marital status as covariates.

I estimated the following $\beta$ coefficients:
`r round(coef(model.6), 4)`.

```{r p2}
# calculate dffits diagnostic
diagnostics <- dffits(model.6)

# inspect influential points
qplot(1:nrow(data), dffits(model.6))

# remove with abs(dffits) > 0.25
influential <- which(abs(diagnostics) > 0.25)

data.no.inf <- data %>%
    filter(!row_number() %in% influential)

# refit model
model.6.refit <- glm(large.exp~lc5 + chd5 + race + 
                     male * age + factor(marital),
                     family="binomial", data=data.no.inf)
```

If we drop these somewhat influential points, we see $\beta$ coefficients
of `r round(coef(model.6.refit), 4)`.

For the most part, the coefficients did not change _too much_, so I
decided to keep these points for the rest of the analysis.

Problem 3
=========
```{r p3}
# all data
data <- data %>%
    mutate(prob=predict(model.6, newdata=data, type=c("response")))

roc <- roc(large.exp~prob, data=data)
auc <- auc(roc)
plot(roc, legacy.axes=TRUE)

main.effects.model <- glm(large.exp ~ lc5 + chd5 + age + male,
                          family="binomial", data=data)
data <- data %>%
    mutate(prob.me=predict(main.effects.model, 
                           newdata=data, type=c("response")))

auc.me <- auc(roc(large.exp~prob.me, data=data))

# cross validation
data <- data %>%
    mutate(id=1:nrow(data))

train <- data %>%
    sample_frac(0.5)

test <- setdiff(data, train)

model.final <- glm(large.exp~lc5 + chd5 + race + 
                   male * age + factor(marital),
                   family="binomial", data=train)
model.main <- glm(large.exp ~ lc5 + chd5 + age + male,
                  family="binomial", data=train)
train <- train %>%
    mutate(prob=predict(model.final, newdata=train, type=c("response")),
           prob.me=predict(model.main, newdata=train, type=c("response")))

test <- test %>%
    mutate(prob=predict(model.final, newdata=test, type=c("response")),
           prob.me=predict(model.main, newdata=test, type=c("response")))

auc.train <- auc(roc(large.exp~prob, data=train))
auc.test <- auc(roc(large.exp~prob, data=test))
auc.train.me <- auc(roc(large.exp~prob.me, data=train))
auc.test.me <- auc(roc(large.exp~prob.me, data=test))
```

Using my final model, I calculated an AUC of `r auc` and with the main
effects model `r auc.me`.  
When I performed cross validation, I saw auc
of `r auc.train` on the training data and `r auc.test` on the test data.  
For the main effects model, I saw an auc of `r auc.train.me` and 
`r auc.test.me` on the training and test data respectively.

Problem 4
=========
In order to predict whether or not someone will have a large expenditure
(defined as a medical expenditure more than 20,000 dollars), we fit
a logistic regression model to our dataset. We found that this model
performs well, predicting that someone with a large expenditure has
a higher probability of this large expenditure than someone without
a large expenditure `r round(auc.test * 100, 2)`% of the time in a cross
validation. However we decided on variables to keep in the model using 
the full model. Future work should base these decisions only on training
data.

Part 2
======
```{r part2, echo=FALSE}
library(survival)

data <- read.table("endometrial-cancer-breslow-day.txt") %>%
    tbl_df() %>%
    rename(set=V1, case=V2, age=V3, age.group=V4, estrogen=V5,
           gall.disease=V6, hypertension=V7, obesity=V8, non.est=V9) %>%
    # code disease to 0/1
    mutate(estrogen=estrogen - 1,
           gall.disease = gall.disease - 1,
           hypertension = hypertension - 1,
           obesity = obesity - 1,
           non.est = non.est - 1,
           # make obesity a factor
           obesity=factor(obesity),
           # id variable
           id=1:n())
    
# 1-1 design
data.1 <- data %>%
    group_by(set) %>%
    filter(id <= min(id) + 1) 

model.1.a <- clogit(case ~ estrogen + gall.disease + 
                    hypertension + strata(set), data=data.1)
model.1.b <- clogit(case ~ estrogen + gall.disease + hypertension +
                    (obesity) + strata(set), data=data.1)
model.1.c <- clogit(case ~ estrogen * gall.disease + hypertension *
                    (age) + strata(set), data=data.1)
model.1.d <- clogit(case ~ estrogen * gall.disease + hypertension +
                    obesity + strata(set), data=data.1)
model.1.e <- clogit(case ~ estrogen * gall.disease + hypertension *
                    obesity + strata(set), data=data.1)

# LRTs
lrt <- function(null, ext) {
    d.1 <- -2 * null$loglik[2]
    d.2 <- -2 * ext$loglik[2]
    df <- length(ext$coef) - length(null$coef)
    1 - pchisq(d.1 - d.2, df)
}

a.b <- lrt(model.1.a, model.1.b)
a.c <- lrt(model.1.a, model.1.c)
a.d <- lrt(model.1.a, model.1.d)
a.e <- lrt(model.1.a, model.1.e)

# 1 - 4 design
model.4.a <- clogit(case ~ estrogen + gall.disease + 
                    hypertension + strata(set), data=data)
model.4.b <- clogit(case ~ estrogen + gall.disease + hypertension +
                    (obesity) + strata(set), data=data)
model.4.c <- clogit(case ~ estrogen * gall.disease + hypertension *
                    (age) + strata(set), data=data)
model.4.d <- clogit(case ~ estrogen * gall.disease + hypertension +
                    obesity + strata(set), data=data)
model.4.e <- clogit(case ~ estrogen * gall.disease + hypertension *
                    obesity + strata(set), data=data)

a.b <- lrt(model.4.a, model.4.b)
a.c <- lrt(model.4.a, model.4.c)
a.d <- lrt(model.4.a, model.4.d)
d.e <- lrt(model.4.d, model.4.e)

```

\newpage
Using a matched case control study to study the effect of estrogen use on endometrial cancer. The data contained 63 matched sets, where cases were matched to 4 controls by being alive in the same community at the time of the case was diagnosed, having age within 1 year, same marital status and entering the community at roughly the same time.

```{r fig1, echo=FALSE}
theme_set(theme_classic())

est <- data %>%
    group_by(set) %>%
    summarise(estrogen.users=sum(estrogen))

data <- data %>%
    left_join(est, by="set")

ggplot(data, aes(x=reorder(set, -age, median), y=age)) +
    geom_boxplot(aes(colour=factor(estrogen.users))) +
    xlab("") +
    ylab("Age") +
    ggtitle("Figure 1") +
    scale_x_discrete(breaks=NULL) +
    labs(colour="Estrogen Users")
```

As is shown above, and is expected from the description of the data, the age of the participants in a group
appear to be within one year of each other. Additionally, the data seem to be evenly distributed by age, and 
the number of estrogen users within a group also seems evenly distributed by age.


```{r fig2, echo=FALSE}
theme_set(theme_classic())

ggplot(data, aes(x=age, y=case)) +
    geom_jitter(aes(colour=factor(estrogen))) +
    ggtitle("Figure 2")
```

\newpage
For our final model, we adjusted for gall bladder disease, hypertension, 
obesity (categorized as not obsese, obese, or unknown) and the
interaction of estrogen and gall bladder disease. We chose this model
using the likelihood ratio test. Below are the estimates for this model,
using the full data.

```{r tbl1, echo=FALSE}
var <- c("Obesity", "\\hspace*{1em}Not Obese", "\\hspace*{1em}Obese", 
         "\\hspace*{1em}Unknown",
         "Estrogen and Gallbladder Disease", 
         "\\hspace*{1em}Neither", "\\hspace*{1em}Estrogen Only", 
         "\\hspace*{1em}Gallbladder Disease Only", "\\hspace*{1em}Both",
         "Hypertension")

coef.4 <- coef(model.4.d)
coef.4[6] <- sum(coef.4[c(1:2, 6)])
coef.4 <- exp(coef.4)
ci.4 <- confint(model.4.d)
ci.4[6, 1] <- sum(ci.4[c(1:2, 6), 1])
ci.4[6, 2] <- sum(ci.4[c(1:2, 6), 2])
ci.4 <- exp(ci.4)
or.all.4 <- character(length(coef.4))

for (i in 1:length(coef.4)) {
    or.all.4[i] <- sprintf("%3.2f (%3.2f, %3.2f)", 
                           coef.4[i], ci.4[i, 1], ci.4[i, 2])
}

or.4 <- c("", "Reference", or.all.4[4], or.all.4[5],
          "", "Reference", or.all.4[1], or.all.4[2], or.all.4[6],
          or.all.4[3])

tbl.4 <- data.frame(var=var, or=or.4, stringsAsFactors=FALSE)
knitr::kable(tbl.4, col.names=c("Variable", "Odds Ratio"),
             caption="1-4 Matching")
```

We also experimented with using a case and the first control matched to
the case in each set. The point estimates were fairly close, however the
confidence intervals were much wider.

```{r tbl2, echo=FALSE}
coef.1 <- coef(model.1.d)
coef.1[6] <- sum(coef.1[c(1:2, 6)])
coef.1 <- exp(coef.1)
ci.1 <- confint(model.1.d)
ci.1[6, 1] <- sum(ci.1[c(1:2, 6), 1])
ci.1[6, 2] <- sum(ci.1[c(1:2, 6), 2])
ci.1 <- exp(ci.1)
or.all.1 <- character(length(coef.1))

for (i in 1:length(coef.1)) {
    or.all.1[i] <- sprintf("%3.2f (%3.2f, %3.2f)", 
                           coef.1[i], ci.1[i, 1], ci.1[i, 2])
}

or.1 <- c("", "Reference", or.all.1[4], or.all.1[5],
          "", "Reference", or.all.1[1], or.all.1[2], or.all.1[6],
          or.all.1[3])

tbl.1 <- data.frame(var=var, or=or.1, stringsAsFactors=FALSE)
knitr::kable(tbl.1, col.names=c("Variable", "Odds Ratio"),
             caption="1-1 Matching")
```

In either case, we found that estrogen use increased the odds of 
endometrial cancer. However, this associated was only statistically
significant in the 1-4 matching.

\newpage

Code used for problem 2
=======================

```{r, eval=FALSE} 
library(survival)

data <- read.table("endometrial-cancer-breslow-day.txt") %>%
    tbl_df() %>%
    rename(set=V1, case=V2, age=V3, age.group=V4, estrogen=V5,
           gall.disease=V6, hypertension=V7, obesity=V8, non.est=V9) %>%
    # code disease to 0/1
    mutate(estrogen=estrogen - 1,
           gall.disease = gall.disease - 1,
           hypertension = hypertension - 1,
           obesity = obesity - 1,
           non.est = non.est - 1,
           # make obesity a factor
           obesity=factor(obesity),
           # id variable
           id=1:n())
    
# 1-1 design
data.1 <- data %>%
    group_by(set) %>%
    filter(id <= min(id) + 1) 

model.1.a <- clogit(case ~ estrogen + gall.disease + 
                    hypertension + strata(set), data=data.1)
model.1.b <- clogit(case ~ estrogen + gall.disease + hypertension +
                    (obesity) + strata(set), data=data.1)
model.1.c <- clogit(case ~ estrogen * gall.disease + hypertension *
                    (age) + strata(set), data=data.1)
model.1.d <- clogit(case ~ estrogen * gall.disease + hypertension +
                    obesity + strata(set), data=data.1)
model.1.e <- clogit(case ~ estrogen * gall.disease + hypertension *
                    obesity + strata(set), data=data.1)

# LRTs
lrt <- function(null, ext) {
    d.1 <- -2 * null$loglik[2]
    d.2 <- -2 * ext$loglik[2]
    df <- length(ext$coef) - length(null$coef)
    1 - pchisq(d.1 - d.2, df)
}

a.b <- lrt(model.1.a, model.1.b)
a.c <- lrt(model.1.a, model.1.c)
a.d <- lrt(model.1.a, model.1.d)
a.e <- lrt(model.1.a, model.1.e)

# 1 - 4 design
model.4.a <- clogit(case ~ estrogen + gall.disease + 
                    hypertension + strata(set), data=data)
model.4.b <- clogit(case ~ estrogen + gall.disease + hypertension +
                    (obesity) + strata(set), data=data)
model.4.c <- clogit(case ~ estrogen * gall.disease + hypertension *
                    (age) + strata(set), data=data)
model.4.d <- clogit(case ~ estrogen * gall.disease + hypertension +
                    obesity + strata(set), data=data)
model.4.e <- clogit(case ~ estrogen * gall.disease + hypertension *
                    obesity + strata(set), data=data)

a.b <- lrt(model.4.a, model.4.b)
a.c <- lrt(model.4.a, model.4.c)
a.d <- lrt(model.4.a, model.4.d)
d.e <- lrt(model.4.d, model.4.e)

theme_set(theme_classic())

est <- data %>%
    group_by(set) %>%
    summarise(estrogen.users=sum(estrogen))

data <- data %>%
    left_join(est, by="set")

ggplot(data, aes(x=reorder(set, -age, median), y=age)) +
    geom_boxplot(aes(colour=factor(estrogen.users))) +
    xlab("") +
    ylab("Age") +
    ggtitle("Figure 1") +
    scale_x_discrete(breaks=NULL) +
    labs(colour="Estrogen Users")

theme_set(theme_classic())

ggplot(data, aes(x=age, y=case)) +
    geom_jitter(aes(colour=factor(estrogen))) +
    ggtitle("Figure 2")

var <- c("Obesity", "\\hspace*{1em}Not Obese", "\\hspace*{1em}Obese", 
         "\\hspace*{1em}Unknown",
         "Estrogen and Gallbladder Disease", 
         "\\hspace*{1em}Neither", "\\hspace*{1em}Estrogen Only", 
         "\\hspace*{1em}Gallbladder Disease Only", "\\hspace*{1em}Both")

coef.4 <- coef(model.4.d)
coef.4[6] <- sum(coef.4[c(1:2, 6)])
coef.4 <- exp(coef.4)
ci.4 <- confint(model.4.d)
ci.4[6, 1] <- sum(ci.4[c(1:2, 6), 1])
ci.4[6, 2] <- sum(ci.4[c(1:2, 6), 2])
ci.4 <- exp(ci.4)
or.all.4 <- character(length(coef.4))

for (i in 1:length(coef.4)) {
    or.all.4[i] <- sprintf("%3.2f (%3.2f, %3.2f)", 
                           coef.4[i], ci.4[i, 1], ci.4[i, 2])
}

or.4 <- c("", "Reference", or.all.4[4], or.all.4[5],
          "", "Reference", or.all.4[1], or.all.4[2], or.all.4[6])

tbl.4 <- data.frame(var=var, or=or.4, stringsAsFactors=FALSE)
knitr::kable(tbl.4, col.names=c("Variable", "Odds Ratio"),
             caption="1-4 Matching")

coef.1 <- coef(model.1.d)
coef.1[6] <- sum(coef.1[c(1:2, 6)])
coef.1 <- exp(coef.1)
ci.1 <- confint(model.1.d)
ci.1[6, 1] <- sum(ci.1[c(1:2, 6), 1])
ci.1[6, 2] <- sum(ci.1[c(1:2, 6), 2])
ci.1 <- exp(ci.1)
or.all.1 <- character(length(coef.1))

for (i in 1:length(coef.1)) {
    or.all.1[i] <- sprintf("%3.2f (%3.2f, %3.2f)", 
                           coef.1[i], ci.1[i, 1], ci.1[i, 2])
}

or.1 <- c("", "Reference", or.all.1[4], or.all.1[5],
          "", "Reference", or.all.1[1], or.all.1[2], or.all.1[6])

tbl.1 <- data.frame(var=var, or=or.1, stringsAsFactors=FALSE)
knitr::kable(tbl.1, col.names=c("Variable", "Odds Ratio"),
             caption="1-1 Matching")
coef.1 <- coef(model.1.d)
coef.1[6] <- sum(coef.1[c(1:2, 6)])
coef.1 <- exp(coef.1)
ci.1 <- confint(model.1.d)
ci.1[6, 1] <- sum(ci.1[c(1:2, 6), 1])
ci.1[6, 2] <- sum(ci.1[c(1:2, 6), 2])
ci.1 <- exp(ci.1)
or.all.1 <- character(length(coef.1))

for (i in 1:length(coef.1)) {
    or.all.1[i] <- sprintf("%3.2f (%3.2f, %3.2f)", 
                           coef.1[i], ci.1[i, 1], ci.1[i, 2])
}

or.1 <- c("", "Reference", or.all.1[4], or.all.1[5],
          "", "Reference", or.all.1[1], or.all.1[2], or.all.1[6])

tbl.1 <- data.frame(var=var, or=or.1, stringsAsFactors=FALSE)
knitr::kable(tbl.1, col.names=c("Variable", "Odds Ratio"),
             caption="1-1 Matching")
```
