---
title:  Homework 2
author: "[Jacob Carey](mailto:jcarey15@jhu.edu)"
date:   "`r Sys.Date()`"
output: pdf_document
---

```{r chunks, echo=FALSE}
knitr::opts_chunk$set(cache=TRUE,
                      fig.width=6,
                      fig.height=4)
```


```{r readin, echo=FALSE}
load("../data/nmes.rdata")
library(ggplot2)
library(dplyr)
library(tidyr)
library(pROC)
theme_set(theme_classic())
```

Problem 1
=========
```{r p1}
data <- nmes %>%
    filter(lastage >= 40) %>%
    mutate(large.exp=as.integer(totalexp > 2e4)) %>%
    select(large.exp, lc5, chd5, educate, race=RACE3,
           male, age=lastage, region=sregion, marital,
           poverty=povstalb)

model.1 <- glm(large.exp~lc5 + chd5, family="binomial", data=data)
model.2 <- glm(large.exp~lc5 + chd5 + race + male, 
               family="binomial", data=data)
model.3 <- glm(large.exp~lc5 + chd5 + race + male + age, 
               family="binomial", data=data)
model.4 <- glm(large.exp~lc5 + chd5 + race + male * age, 
               family="binomial", data=data)
model.5 <- glm(large.exp~lc5 + chd5 + race + male * age + factor(poverty),
               family="binomial", data=data)
model.6 <- glm(large.exp~lc5 + chd5 + race + male * age + factor(marital),
               family="binomial", data=data)

p.value.1.2 <- 1 - pchisq(model.1$deviance - model.2$deviance,
                          model.1$df.residual - model.2$df.residual)
p.value.1.3 <- 1 - pchisq(model.1$deviance - model.3$deviance,
                          model.1$df.residual - model.3$df.residual)
p.value.3.4 <- 1 - pchisq(model.3$deviance - model.4$deviance,
                          model.3$df.residual - model.4$df.residual)
p.value.4.5 <- 1 - pchisq(model.4$deviance - model.5$deviance,
                          model.4$df.residual - model.5$df.residual)
p.value.4.6 <- 1 - pchisq(model.4$deviance - model.6$deviance,
                          model.4$df.residual - model.6$df.residual)
```

Using the likelihood ratio test we calculate the following p values:  
Comparing model 1 to 2: `r round(p.value.1.2, 4)`  
Comparing model 1 to 3: `r round(p.value.1.3, 4)`  
Comparing model 3 to 4: `r round(p.value.3.4, 4)`  
Comparing model 4 to 5: `r round(p.value.4.5, 4)`  
Comparing model 4 to 6: `r round(p.value.4.6, 4)`  

Problem 2
=========
Based on these calculations, I decided to use model 6, which include
lc5, chd5, race, male, age, an interaction between male and age, and
marital status as covariates.

I estimated the following $\beta$ coefficients:
`r round(coef(model.6), 4)`.

```{r p2}
# calculate dffits diagnostic
diagnostics <- dffits(model.6)

# inspect influential points
qplot(1:nrow(data), dffits(model.6))

# remove with abs(dffits) > 0.25
influential <- which(abs(diagnostics) > 0.25)

data.no.inf <- data %>%
    filter(!row_number() %in% influential)

# refit model
model.6.refit <- glm(large.exp~lc5 + chd5 + race + 
                     male * age + factor(marital),
                     family="binomial", data=data.no.inf)
```

If we drop these somewhat influential points, we see $\beta$ coefficients
of `r round(coef(model.6.refit), 4)`.

For the most part, the coefficients did not change _too much_, so I
decided to keep these points for the rest of the analysis.

Problem 3
=========
```{r p3}
# all data
data <- data %>%
    mutate(prob=predict(model.6, newdata=data, type=c("response")))

roc <- roc(large.exp~prob, data=data)
auc <- auc(roc)
plot(roc, legacy.axes=TRUE)

main.effects.model <- glm(large.exp ~ lc5 + chd5 + age + male,
                          family="binomial", data=data)
data <- data %>%
    mutate(prob.me=predict(main.effects.model, 
                           newdata=data, type=c("response")))

auc.me <- auc(roc(large.exp~prob.me, data=data))

# cross validation
data <- data %>%
    mutate(id=1:nrow(data))

train <- data %>%
    sample_frac(0.5)

test <- setdiff(data, train)

model.final <- glm(large.exp~lc5 + chd5 + race + 
                   male * age + factor(marital),
                   family="binomial", data=train)
model.main <- glm(large.exp ~ lc5 + chd5 + age + male,
                  family="binomial", data=train)
train <- train %>%
    mutate(prob=predict(model.final, newdata=train, type=c("response")),
           prob.me=predict(model.main, newdata=train, type=c("response")))

test <- test %>%
    mutate(prob=predict(model.final, newdata=test, type=c("response")),
           prob.me=predict(model.main, newdata=test, type=c("response")))

auc.train <- auc(roc(large.exp~prob, data=train))
auc.test <- auc(roc(large.exp~prob, data=test))
auc.train.me <- auc(roc(large.exp~prob.me, data=train))
auc.test.me <- auc(roc(large.exp~prob.me, data=test))
```

Using my final model, I calculated an AUC of `r auc` and with the main
effects model `r auc.me`.  
When I performed cross validation, I saw auc
of `r auc.train` on the training data and `r auc.test` on the test data.  
For the main effects model, I saw an auc of `r auc.train.me` and 
`r auc.test.me` on the training and test data respectively.

Problem 4
=========
In order to predict whether or not someone will have a large expenditure
(defined as a medical expenditure more than 20,000 dollars), we fit
a logistic regression model to our dataset. We found that this model
performs well, predicting that someone with a large expenditure has
a higher probability of this large expenditure than someone without
a large expenditure `r round(auc.test * 100, 2)`% of the time in a cross
validation. However we decided on variables to keep in the model using 
the full model. Future work should base these decisions only on training
data.

Part 2
======
```{r part2, echo=FALSE}
library(survival)
data <- read.table("endometrial-cancer-breslow-day.txt") %>%
    tbl_df() %>%
    rename(set=V1, case=V2, age=V3, age.group=V4, estrogen=V5,
           gall.disease=V6, hypertension=V7, obesity=V8, non.est=V9) %>%
    # code disease to 0/1
    mutate(estrogen=estrogen - 1,
           gall.disease = gall.disease - 1,
           hypertension = hypertension - 1,
           obesity = obesity - 1,
           non.est = non.est - 1,
           # make obesity a factor
           obesity=factor(obesity),
           # id variable
           id=1:n())
    
# 1-1 design
data.1 <- data %>%
    group_by(set) %>%
    filter(id <= min(id) + 1) 

model.1.a <- clogit(case ~ estrogen + gall.disease + 
                    hypertension + strata(set), data=data.1)
model.1.b <- clogit(case ~ (estrogen + gall.disease + hypertension) *
                    (age) + strata(set), data=data.1)
model.1.c <- clogit(case ~ estrogen + gall.disease + hypertension *
                    (obesity) + strata(set), data=data.1)
model.1.d <- clogit(case ~ estrogen * gall.disease + hypertension *
                    (age) + strata(set), data=data.1)
model.1.e <- clogit(case ~ estrogen + gall.disease * 
                      hypertension + strata(set), data=data.1)

# 1 - 4 design
model.4.a <- clogit(case ~ estrogen + gall.disease + 
                    hypertension + strata(set), data=data)
model.4.b <- clogit(case ~ (estrogen + gall.disease + hypertension) *
                    (age) + strata(set), data=data)
model.4.c <- clogit(case ~ estrogen + gall.disease + hypertension *
                    (obesity) + strata(set), data=data)
model.4.d <- clogit(case ~ estrogen * gall.disease + hypertension *
                    (age) + strata(set), data=data)
model.4.e <- clogit(case ~ estrogen + gall.disease * 
                      hypertension + strata(set), data=data)
# summary(model.4.a)$logtest
# summary(model.4.b)$logtest
# summary(model.4.c)$logtest
# summary(model.4.d)$logtest
# summary(model.4.e)$logtest
```

\newpage
Using a matched case control study to study the effect of estrogen use on endometrial cancer. The data contained 63 matched sets, where cases were matched to 4 controls by being alive in the same community at the time of the case was diagnosed, having age within 1 year, same marital status and entering the community at roughly the same time.

```{r fig1, echo=FALSE}
est <- data %>%
    group_by(set) %>%
    summarise(estrogen.users=sum(estrogen))

data <- data %>%
    left_join(est, by="set")

ggplot(data, aes(x=reorder(set, -age, median), y=age)) +
    geom_boxplot(aes(colour=factor(estrogen.users))) +
    xlab("") +
    ylab("Age") +
    ggtitle("Figure 1") +
    scale_x_discrete(breaks=NULL) +
#     theme_classic() +
    labs(colour="Estrogen Users")
```

As is shown above, and is expected from the description of the data, the age of the participants in a group
appear to be within one year of each other. Additionally, the data seem to be evenly distributed by age, and 
the number of estrogen users within a group also seems evenly distributed by age.

```{r fig2, echo=FALSE}
ggplot(data, aes(x=age, y=case)) +
    geom_jitter(aes(colour=factor(estrogen))) +
    ggtitle("Figure 2")
#     theme_classic()
```
